---
title: "How Long Prime Ministers of Canada Lived"
author: 
  - Zijun Meng
thanks: "Code and data are available at: https://github.com/FrankMengZJ/PM_CA"
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
```

In this project, I examined how long the Prime Ministers of Canada lived. I use @citeR, @xml2 and @rvest to get data from @wiki, then used @tidyverse and @janitor to cleaned the data, and finally used @ggplot2 to create a graph of the data.

@fig-fm provides a visual overview of the lifespans of Canadian Prime Ministers. Red bars indicate those who have passed away, while blue bars represent the ones still living. The varying lengths of the bars span from each Prime Minister's year of birth, with blue bars reaching up to the present, showcasing the current ages of living leaders.

A quick glance at the chart reveals that most Prime Ministers were born in the 19th or 20th centuries, in line with Canada's history as a self-governing dominion since 1867. The chart does not explicitly show a trend in lifespans over time but does highlight that most recent leaders, born in the later part of the 20th century, are still alive, possibly reflecting modern advancements in health and well-being.

```{r}
#| label: fig-pm
#| fig-cap: How long each prime minister of Canada lived
#| echo: false

cleaned_list<- read.csv(file=here("data/analysis_data/analysis_data.csv"))
cleaned_list |>
  mutate(
    still_alive = if_else(is.na(died), "Yes", "No"),
    died = if_else(is.na(died), as.integer(2024), died)
  ) |>
  mutate(name = as_factor(name)) |>
  ggplot(
    aes(x = birth, xend = died, y = name, yend = name, color = still_alive)
  ) +
  geom_segment() +
  labs(
    x = "Year of birth", y = "Prime minister", color = "PM is currently alive"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

The data source is from @wiki which provided a list of prime ministers of Canada, including their names, birth, death and other information which is not related to this topic. Using @rvest package, which is designed for web scraping, the code first reads the HTML content of the webpage. It then uses selectors to pinpoint the specific table that contains the information on Prime Ministers.

@tidyverse comes into play for cleaning and organizing the data. The clean_names() function from  @janitor  standardizes the column names for easier reference. The select() function helps to isolate the column with the names and birth-death dates of the Prime Ministers, and distinct() ensures that only unique entries are retained, removing any potential duplicates.

The data is then further processed to separate the names from the birth and death dates. Regular expressions are used to extract the years, and any missing birth years are filled in with the corresponding birth year data from entries that only have a birth year available, which indicates that the person is still alive.

Finally, @ggplot2  is used to create the horizontal bar chart visualizing the lifespans of the Prime Ministers. 

I encountered several problems when working the data. There were several tables in @wiki, at first I used html_element() function which only returns the first element it finds, so I searched online and found out I should use html_elements() instead.

Another problem I met was when I was creating the graph, the birth and death time wasn't integers, I realized that when cleaning the data, I just separated the strings without further adjustments. I transferred the strings into integers and the problem was solved. Next time, I will pay attention to types of data when doing data cleaning.
\newpage
# References


